Real-space computation of $E$/$B$-mode maps I: Formalism and Compact Kernels

### 2017-11-22 articles/formalism/sections/real_space_formalism.tex
added explanation for the alpha and gamma as being the radiating and the convolution kernels, still sharpening the explanation

&=& -0.5 \yzmat{i} \qutoxd \ymatc{j} \qutox   \,, \\
&=& -0.5 \begin{bmatrix} \sum ({}_{0}Y_i ~{}_{2}Y^{T*}_j  +  {}_{0}Y_i~ {}_{-2}Y^{T*}_j) & {\rm i}  \sum ({}_{0}Y_i ~ {}_{2}Y^{T*}_j - {}_{0}Y_i ~{}_{-2}Y^{T*}_j)  \\  - {\rm i} \sum  ({}_{0}Y_i ~ {}_{2}Y^{T*}_j - {}_{0}Y_i~ {}_{-2}Y^{T*}_j) & \sum ({}_{0}Y_i~ {}_{2}Y^{T*}_j + {}_{0}Y_i ~{}_{-2}Y^{T*}_j)  \end{bmatrix} \,, \label{eq:qu2eb_ker_1}

@@

where the symbol ${}_{0}Y_i$ is used to denote the sub-matrix ${}_{0}Y_{\hat{n}_i \times \ell m} \equiv {}_{0}Y_{\ell m}(\hat{n}_i)$, the symbol ${}_{\pm 2}Y^{T*}_j$ is used to denote the matrix ${}_{\pm 2}Y^*_{\ell m \times \hat{n}_j} \equiv {}_{\pm 2}Y^*_{\ell m}(\hat{n}_j)$ and the summation is over the multipole indices $\ell,m$. Using the conjugation properties of the spin spherical harmonic functions it can be shown that the following relation holds true,

@@

 \left [\sum_{\ell m} {}_{0}Y_{\ell m}(\hat{n}_i){}_{+2}Y^*_{\ell m}(\hat{n}_j)\right]^* = \sum_{\ell m} {}_{0}Y_{\ell m}(\hat{n}_i){}_{-2}Y^*_{\ell m}(\hat{n}_j) \,.

@@

 \sum_{m}{{}_{s_1}Y}^*_{\ell m}(\hat{n}_i)\,{{}_{s_2}Y}_{\ell m}(\hat{n}_j) = \sqrt{\frac{2\ell+1}{4 \pi}} {{}_{s_2}}Y_{\ell \,-s_1}(\beta,\alpha) e^{- i s_2 \gamma} \,,

@@

where $\alpha, ~\beta ~\&~ \gamma$ denote the Euler angles. Therefore the different parts of the real space operator $\bar{O}$  are completely specified in terms of the complex function,

@@

\mathcal{M}( \hat{n}_i, \hat{n}_j)  &=& \mathcal{M}_{r} + i \mathcal{M}_{i}  \,,\nonumber \\ 
&=&\sum_{\ell m} {{_0}Y}_{\ell m}(\hat n_i) \, {{_{-2}}Y}^*_{\ell m}(\hat n_j) = \sum_{\ell} \sqrt{\frac{2\ell+1}{ 4 \pi}}{{_0Y}_{\ell 2}}(\beta_{ij},\alpha_{ij})\,,\\
&=&  \Big [ \cos(2 \alpha_{ij}) + i \sin(2 \alpha_{ij} ) \Big]   \sum_{\ell=\ell_{\rm min}}^{\ell_{\rm max}} {\frac{2\ell+1}{ 4 \pi}} \sqrt{\frac{(\ell-2)!}{(\ell+2)!}}P_{\ell 2} (\cos\beta_{ij}) \,, \label{eq:rad_ker_queb} \\
&=&  \Big [ \cos(2 \alpha_{ij}) + i \sin(2 \alpha_{ij} ) \Big] {{}_{\mm}f}(\beta_{ij},\ell_{\rm min},\ell_{\rm max}) \,, 

@@

\bar O =-\bmat  \mathcal{M}_{r} & \mathcal{M}_{i} \\  -\mathcal{M}_{i}  & \mathcal{M}_{r} \emat_{2 N_{\rm pix} \times 2 N_{pix}} = -{{}_{\mm}f}(\beta_{ij},\ell_{\rm min},\ell_{\rm max})\bmat \cos(2 \alpha_{ij}) & \sin(2\alpha_{ij})\\  -\sin(2 \alpha_{ij})  & \cos(2 \alpha_{ij}) \emat \,,

@@

where indices i,j map to the location $\hat{n}_i$ and $\hat{n}_j$ on the sphere. \revisit{A similar equation for real space $E$ \& $B$ operators was derived in \cite{Zaldarriaga2001a}, however those results were derived for the flat sky case and did not explicitly derive the radial kernel.} \comment{A discussion on this should be in the conclusions.}

The scalar fields $E$ \& $B$ can now be directly derived from the measured Stokes $Q$ \& $U$ parameters by evaluating the following convolution,
%
\beq \label{eq:qu2eb_convolution_explicit}
\bmat E_i \\ B_i  \emat= -\Delta \Omega \sum_{j=1}^{N_{\rm pix}}{{}_{\mm}f}(\beta_{ij},\ell_{\rm min},\ell_{\rm max})\bmat \cos(2 \alpha_{ij}) & \sin(2\alpha_{ij})\\  -\sin(2 \alpha_{ij})  & \cos(2 \alpha_{ij}) \emat  \bmat Q_j \\ U_j  \emat \,,
\eeq
%
where $\Delta \Omega$ denotes the pixel area and all the symbols have their usual meaning. 
%\revisit{This has an elegant interpretation: to derive the E and/or B field at any given position we need to find the cosine quadrupole transform and the sine quadrupole transform of the Stokes Q \& U parameters on circles around this position, weigh the transform by the value of the function $f(\beta,\ell_{\rm min},\ell_{\rm max})$, $\beta$ being the radius of the circle and sum up the results with appropriate signs, to construct the respective scalar fields.} 
The above equation can be expressed more concisely as follows,
%
\begin{subequations} \label{eq:qu2eb_simple}
\beqry 
[E + iB](\hat{n}_0) &=& - \Delta \Omega \sum_{j=1}^{N_{\rm pix}} \left(\sum_{\ell=\ell_{\rm min}}^{\ell_{\rm max}} \frac{2 \ell+1}{4 \pi} \sqrt{\frac{(\ell-2)!}{(\ell+2)!}} P_{\ell}^{2}(\beta_{0j}) \right) {\Bigg( e^{-i2 \alpha_{0j}}   {}_{+2}X (\hat{n}_j) \Bigg)} \,, \label{eq:qu2eb_physical}\\
&=& - \left( \left[ \sum_{\ell=\ell_{\rm min}}^{\ell_{\rm max}} \sqrt{\frac{2 \ell+1}{4 \pi}}Y^*_{\ell 2} \right]  \circ {}_{+2}X \right) \,, \\
&=& - \Bigg\lbrace \mathcal{M}^* \circ {}_{+2}X \Bigg\rbrace(\hat{n}_0)\,, \label{eq:qu2eb_convolution} 
\eeqry
\end{subequations}
%
where $\circ$ denotes a convolution and the spherical harmonic functions denote the rotated functions such that the pole of the function coincides with the direction $\hat{n}_0$ and the reference zero for the azimuthal angle is the local longitude $\phi_0$.

>>>>

&=& -0.5 \yzmat{e} \qutoxd \ymatc{q} \qutox   \,, \\
&=& -0.5 \begin{bmatrix} \sum ({}_{0}Y_e ~{}_{2}Y^{T*}_q  +  {}_{0}Y_e~ {}_{-2}Y^{T*}_q) & {\rm i}  \sum ({}_{0}Y_e ~ {}_{2}Y^{T*}_q - {}_{0}Y_e ~{}_{-2}Y^{T*}_q)  \\  - {\rm i} \sum  ({}_{0}Y_e ~ {}_{2}Y^{T*}_q - {}_{0}Y_e~ {}_{-2}Y^{T*}_q) & \sum ({}_{0}Y_e~ {}_{2}Y^{T*}_q + {}_{0}Y_e ~{}_{-2}Y^{T*}_q)  \end{bmatrix} \,, \label{eq:qu2eb_ker_1}

@@

where the symbol ${}_{0}Y_e$ is used to denote the sub-matrix ${}_{0}Y_{\hat{n}_e \times \ell m} \equiv {}_{0}Y_{\ell m}(\hat{n}_e)$, the symbol ${}_{\pm 2}Y^{T*}_q$ is used to denote the matrix ${}_{\pm 2}Y^*_{\ell m \times \hat{n}_q} \equiv {}_{\pm 2}Y^*_{\ell m}(\hat{n}_q)$ and the summation is over the multipole indices $\ell,m$. \revisit{We have purposefully used the index ``e'' to denote the location where the scalar fields are measured and the index ``q'' to denote the location where the Stokes parameters are measured.} Using the conjugation properties of the spin spherical harmonic functions it can be shown that the following relation holds true,

@@

 \left [\sum_{\ell m} {}_{0}Y_{\ell m}(\hat{n}_e){}_{+2}Y^*_{\ell m}(\hat{n}_q)\right]^* = \sum_{\ell m} {}_{0}Y_{\ell m}(\hat{n}_e){}_{-2}Y^*_{\ell m}(\hat{n}_q) \,.

@@

 \sum_{m}{{}_{s_1}Y}^*_{\ell m}(\hat{n}_i)\,{{}_{s_2}Y}_{\ell m}(\hat{n}_j) = \sqrt{\frac{2\ell+1}{4 \pi}} {{}_{s_2}}Y_{\ell \,-s_1}(\beta_{ij},\alpha_{ij}) e^{- i s_2 \gamma_{ij}} \,,

@@

where $\alpha_{ij}, ~\beta_{ij} ~\&~ \gamma_{ij}$ denote the Euler angles. Therefore the different parts of the real space operator $\bar{O}$  are completely specified in terms of the complex function,

@@

\mathcal{M}( \hat{n}_e, \hat{n}_q)  &=& \mathcal{M}_{r} + i \mathcal{M}_{i}  \,,\nonumber \\ 
&=&\sum_{\ell m} {{_0}Y}_{\ell m}(\hat n_e) \, {{_{-2}}Y}^*_{\ell m}(\hat n_q) = \sum_{\ell} \sqrt{\frac{2\ell+1}{ 4 \pi}}{{_0Y}_{\ell 2}}(\beta_{qe},\alpha_{qe})\,,\\
&=&  \Big [ \cos(2 \alpha_{qe}) + i \sin(2 \alpha_{qe} ) \Big]   \sum_{\ell=\ell_{\rm min}}^{\ell_{\rm max}} {\frac{2\ell+1}{ 4 \pi}} \sqrt{\frac{(\ell-2)!}{(\ell+2)!}}P_{\ell 2} (\cos\beta_{qe}) \,, \label{eq:rad_ker_queb} \\
&=&  \Big [ \cos(2 \alpha_{qe}) + i \sin(2 \alpha_{qe} ) \Big] {{}_{\mm}f}(\beta_{qe},\ell_{\rm min},\ell_{\rm max}) \,, 

@@

\bar O =-\bmat  \mathcal{M}_{r} & \mathcal{M}_{i} \\  -\mathcal{M}_{i}  & \mathcal{M}_{r} \emat_{2 N_{\rm pix} \times 2 N_{pix}} = -{{}_{\mm}f}(\beta_{qe},\ell_{\rm min},\ell_{\rm max})\bmat \cos(2 \alpha_{qe}) & \sin(2\alpha_{qe})\\  -\sin(2 \alpha_{qe})  & \cos(2 \alpha_{qe}) \emat \,.

@@

Specifically note that $\alpha_{qe}, ~\beta_{qe} ~\&~ \gamma_{qe}$ denote the Euler angles which rotate the local cartesian system at the location $\hat{n}_q$ where the Stokes field is measured to the cartesian system at the location $\hat{n}_e$ where the scalar fields are evaluated: $\hat{n}_q \xrightarrow{\mathcal{R}(\alpha_{qe},\beta_{qe},\gamma_{qe})} \hat{n}_e$. The Euler angles for the inverse rotation are given by the Euler angles $\alpha_{eq}=-\gamma_{qe}$, $\beta_{eq} = -\beta_{qe}$ and  $\gamma_{eq} =-\alpha_{qe}$\cite{varshalovich}. Since the real space kernels will only depend on the cosine of the Euler angle $\beta$, they are immune to its change of sign. However the ordering of indices  on the Euler angle $\alpha$ and $\gamma$  is important and leads to two different interpretations of the real space kernels.

\textit{Convolution:} Now working from the position of the pixel ``e'' where the scalar modes need to be evaluated, the Euler angles to the surrounding pixels are: $\alpha_{eq}, \beta_{eq},\gamma_{eq}$. One can now think of the value of scalar fields $E$ \& $B$ at the central pixel ``e" as resulting from the convolution over the measured Stokes $Q$ \& $U$ parameters in the surrounding pixels ``q''. This convolution is defined in terms of the Euler angle $\gamma_{eq}$ and is given by the following equation,
%
\beq \label{eq:qu2eb_convolution_explicit}
\bmat E_e \\ B_e  \emat =- \sum_{q=1}^{N_{\rm pix}}{{}_{\mm}f}(\beta_{eq},\ell_{\rm min},\ell_{\rm max})\bmat \cos(2 \gamma_{eq}) & -\sin(2\gamma_{eq})\\  \sin(2 \gamma_{eq})  & \cos(2 \gamma_{eq}) \emat  \bmat Q_q \\ U_q  \emat \Delta \Omega\,,
\eeq
%
where $\Delta \Omega$ denotes the pixel area and all the symbols have their usual meaning. 
%\revisit{This has an elegant interpretation: to derive the E and/or B field at any given position we need to find the cosine quadrupole transform and the sine quadrupole transform of the Stokes Q \& U parameters on circles around this position, weigh the transform by the value of the function $f(\beta,\ell_{\rm min},\ell_{\rm max})$, $\beta$ being the radius of the circle and sum up the results with appropriate signs, to construct the respective scalar fields.} 
The above equation can be expressed more concisely as follows,
%
\begin{subequations} \label{eq:qu2eb_simple}
\beqry 
[E + iB](\hat{n}_e) &=& - \Delta \Omega \sum_{q=1}^{N_{\rm pix}} \left(\sum_{\ell=\ell_{\rm min}}^{\ell_{\rm max}} \frac{2 \ell+1}{4 \pi} \sqrt{\frac{(\ell-2)!}{(\ell+2)!}} P_{\ell}^{2}(\beta_{eq}) \right) {\Bigg( e^{i2 \gamma_{eq}}   {}_{+2}X (\hat{n}_q) \Bigg)} \,, \label{eq:qu2eb_physical}\\
&=& - \left( \left[ \sum_{\ell=\ell_{\rm min}}^{\ell_{\rm max}} \sqrt{\frac{2 \ell+1}{4 \pi}}Y_{\ell 2}(\beta,\gamma) \right]  \circ {}_{+2}X \right) \,, \\
&=& - \Bigg\lbrace \mathcal{M}^* \circ {}_{+2}X \Bigg\rbrace(\hat{n}_0)\,, \label{eq:qu2eb_convolution} 
\eeqry
\end{subequations}
%
where $\circ$ denotes a convolution and the spherical harmonic functions denote the rotated functions such that the pole of the function coincides with the direction $\hat{n}_e$. %and the reference zero for the azimuthal angle is the local longitude $\phi_0$.

\textit{Radiating:} Working from the position ``q'' where the Stokes parameters are known, the Euler angles to the surrounding pixels are: $\alpha_{qe}, \beta_{qe},\gamma_{qe}$. The E \& B mode maps resulting from the Stokes Q \& U parameters specified at location $\hat{n}_q$ is given by the following expression,
%
\beq \label{eq:qu2eb_radiation_explicit}
\bmat E_e \\ B_e  \emat_q =- {{}_{\mm}f}(\beta_{qe},\ell_{\rm min},\ell_{\rm max})\bmat \cos(2 \alpha_{qe}) & \sin(2\alpha_{qe})\\  -\sin(2 \alpha_{qe})  & \cos(2 \alpha_{qe}) \emat  \bmat Q_q \\ U_q  \emat \Delta \Omega \,, 
\eeq
%
and the total map of E \& B modes can be simply evaluated by summing over the contribution from the Stokes parameters at each location $\hat{n}_q$ : $\bar{S} = \sum_{q=1}^{N_{\rm pix}} \bar{S}_q$.
%$\bmat E_e \\ B_e  \emat = \sum_{q=1}^{N_{\rm pix}} \bmat E_e \\ B_e  \emat_q$.
